\Title

### 2013-06-05 synmut.tex
little changes in discussion

Most positions are only transiently variable and SNVs will either fix or
are lost -- at least in small samples. Given a SNV is at a certain
frequency $\nu$, the probability of fixation is higher for beneficial
SNVs than for neutral ones, which in turn fix more frequently than
deleterious ones. Since a neutral mutation, by definition, has no effect
on the fate of the viruses carrying it, the fixation probability of a
neutral SNV at frequency $\nu$ is the frequency itself, i.e.,
$\pfix(\nu) = \nu$; it goes extinct with probability $1-\nu$. For
instance, if a neutral SNV is observed in half of the sequences, it will
fix with a probability of 50\% (see inset in figure
\ref{fig:aftsyn}). The fixation probability of neutral SNVs is
independent of most model assumptions and is only affected if neutral
SNVs are associated preferentially with either high or low fitness
virus.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By analyzing the fate of single nucleotide variants (SNVs) in
longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
against synonymous substitutions in the relatively conserved regions
C2-C4 of the \env{} gene. Computational modeling suggested that these
SNVs have deleterious effects on the order of $0.002$ and that they are
brought to high frequency through linkage to adaptive mutations.
Comparison with biochemical studies of base pairing propensity in RNA
genome of HIV-1 indicated that these mutations are deleterious, at least
in part, because they disrupt RNA secondary structures. Computer models
of RNA folding predict stable hairpins in these regions that
have been suggested to be functional and termed ``insulating
stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
The weak selection against synonymous variants is compatible with the
negative results of \textit{In vitro} replication assays investigating
the fitness effects of small RNA hairpins in HIV-1
\citep{knoepfel_role_2013}: it would take hundreds of cell culture
passages to detect fitness effects of the order of one per mille. The longitudinal data,
however, spans many years and our analysis is able to quantify the
subtle fitness effect of RNA structure within single infections.

The fixation probabilities and the sojourn times of SNVs represent a rich and simple
summary statistics useful to characterize longitudinal sequence data and compare
to models via computer simulations. These statistics are informative
even in absence of a neutral control and thus suited to analyze
properties of synonymous sites. 
We find selection against synonymous substitutions  despite the
fact that the corresponding sites are not conserved in
cross-sectional data. This is consistent with a recent comparative
analysis of SIV and HIV-1 RNA secondary structure using SHAPE assays and
computational methods \citep{pollom_comparison_2013}. While large scale
patterns of RNA structures tend to agree in both viruses, the individual
base pairs forming the structures are almost always
discordant. Even though the molecular architecture of the structure
changes over time, selection seems to maintain these structures, which reduces the
fixation probability and hence the rate of evolution at synonymous
sites. As expected from this argument, the evolutionary rate
at synonymous sites varies greatly along the HIV-1 genome
\citep{mayrose_towards_2007}. This variation can confound estimates of
selection on proteins substantially \citep{ngandu_extensive_2008}. 
The dynamic nature of HIV secondary structure makes finding the 
compensatory mutations that would restore base pairing in the
longitudinal data difficult, since the exact base pairing pattern is
most likely different than in the reference sequence.

Selection against the majority of synonymous substitutions is probably
true across the genome, but we only observe deleterious synonymous SNVs
at high frequency in \shankaregion{} of \env, where they hitch-hike to
high frequencies on nAb escape mutations. Surprisingly, that non-synonymous
mutations display a fixation pattern as if they were neutral;
comp.~\FIG{pfix}. However, nonsynonymous diversity exceeds
synonymous diversity despite the overall much greater constraints on the
amino acid sequence, suggesting that the majority of high frequency SNVs
are escape mutations despite the fact that are often lost again. We suggest that this
paradoxical behavior could be due to escape
mutations that revert after they themselves are recognized by nAbs, and
the competition between different escape mutations within one epitope. 
Both mechanisms reduce the overall fixation probability and could
explain the observed pattern of fixation.

>>>>

Most positions are only transiently variable and SNVs will either fix or
will be lost -- at least in small samples. Given a SNV is at a certain
frequency $\nu$, the probability of fixation is higher for beneficial
SNVs than for neutral ones, which in turn fix more frequently than
deleterious ones. Since a neutral mutation, by definition, has no effect
on the fate of the viruses carrying it, the fixation probability of a
neutral SNV at frequency $\nu$ is the frequency itself, i.e.,
$\pfix(\nu) = \nu$; it goes extinct with probability $1-\nu$. For
instance, if a neutral SNV is observed in half of the sequences, it will
fix with a probability of 50\% (see inset in figure
\ref{fig:aftsyn}). The fixation probability of neutral SNVs is
independent of most model assumptions and is only affected if neutral
SNVs are associated preferentially with either high or low fitness
virus.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By analyzing the fate of single nucleotide variants (SNVs) in
longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
against synonymous substitutions in the relatively conserved regions
C2-C4 of the \env{} gene. Computational modeling suggested that these
SNVs have deleterious effects on the order of $0.002$ and that they are
brought to high frequency through linkage to adaptive mutations.
Comparison with biochemical studies of base pairing propensity in RNA
genome of HIV-1 indicated that these mutations are deleterious, at least
in part, because they disrupt RNA secondary structures. Computer models
of RNA folding predict stable hairpins in these regions that
have been suggested to be functional and termed ``insulating
stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
The weak selection against synonymous variants is compatible with the
negative results of \textit{In vitro} replication assays investigating
the fitness effects of small RNA hairpins in HIV-1
\citep{knoepfel_role_2013}: it would take hundreds of cell culture
passages to detect fitness effects of the order of one per mille. The longitudinal data,
however, spans many years, hence our analysis is able to quantify the
subtle fitness effect of RNA structure within single infections.

The fixation probabilities and the sojourn times of SNVs represent a rich and simple
summary statistics useful to characterize longitudinal sequence data and compare
to models via computer simulations. These statistics are informative
even in absence of a neutral control and thus suited to analyze
properties of synonymous sites. 
We find selection against synonymous substitutions  despite the
fact that the corresponding sites are not conserved in
cross-sectional data. This is consistent with a recent comparative
analysis of SIV and HIV-1 RNA secondary structure using SHAPE assays and
computational methods \citep{pollom_comparison_2013}. While large scale
patterns of RNA structures tend to agree in both viruses, the individual
base pairs forming the structures are almost always
discordant. Even though the molecular architecture of the structure
changes over time, selection seems to maintain these structures, which reduces the
fixation probability and hence the rate of evolution at synonymous
sites. As expected from this argument, the evolutionary rate
at synonymous sites varies greatly along the HIV-1 genome
\citep{mayrose_towards_2007}. This variation can confound estimates of
selection on proteins substantially \citep{ngandu_extensive_2008}. 
The dynamic nature of HIV secondary structure makes finding the 
compensatory mutations that would restore base pairing in the
longitudinal data difficult; in fact, the exact base pairing pattern is
most likely different than in the reference sequence.

Selection against the majority of synonymous substitutions is probably
true across the genome, but we only observe deleterious synonymous SNVs
at high frequency in \shankaregion{} of \env{}, where they hitchhike to
high frequencies on nAb escape mutations. Surprisingly, nonsynonymous
mutations display a fixation pattern as if they were neutral;
see~\FIG{pfix}. However, nonsynonymous diversity exceeds
synonymous diversity despite the overall much greater constraints on the
amino acid sequence, suggesting that the majority of high frequency SNVs
are escape mutations despite the fact that are often lost again. We suggest that this
paradoxical behavior could be due to (i) escape
mutations that revert after they themselves are recognized by nAbs, or (ii)
the competition between different escape mutations within one epitope. 
Both mechanisms reduce the overall fixation probability and could
explain the observed pattern of fixation.
